---
title: "scMappR"
output: html_document
editor_options: 
  chunk_output_type: console
---

https://cran.r-project.org/web/packages/scMappR/vignettes/scMappR_Vignette.html
https://academic.oup.com/nargab/article/3/1/lqab011/6148840#supplementary-data


# Installation:

```{r}
#BiocManager::install("pcaMethods")
#BiocManager::install("GSVA")
#install.packages("scMappR")
library(scMappR)
```


# Bulk RNA-seq input:

1. gene x replicate normalized count matrix

2. a list of DEGs with gene name, p-value and log2 fold change

Will run scMappR for each time point separately, each time comparing mated and virgin females. 
Only for DE genes (not exons).

Input files:

```{r}
cpm = read.delim("../data/cpm_updated.txt", header = T, as.is = T)
cpm = cpm[, -grep("S", colnames(cpm))]
colnames(cpm) = c("A_virgin_0_5", "A_virgin_1", "A_virgin_2", "A_virgin_3", "A_virgin_4", "A_virgin_5", "A_virgin_6",
                  "A_virgin_8", "A_virgin_12", "A_virgin_24", "A_mated_0_5", "A_mated_1", "A_mated_2", "A_mated_3", 
                  "A_mated_4", "A_mated_5", "A_mated_6", "A_mated_8", "A_mated_12", "A_mated_24", "B_virgin_0_5",
                  "B_virgin_1", "B_virgin_2", "B_virgin_3", "B_virgin_4", "B_virgin_5", "B_virgin_6", "B_virgin_8",
                  "B_virgin_12", "B_virgin_24", "B_mated_0_5", "B_mated_1", "B_mated_2", "B_mated_3", "B_mated_4", "B_mated_5",
                  "B_mated_6", "B_mated_8", "B_mated_12", "B_mated_24", "C_virgin_0_5", "C_virgin_1", "C_virgin_2",
                  "C_virgin_3", "C_virgin_4", "C_virgin_5", "C_virgin_8", "C_virgin_12", "C_virgin_24", "C_mated_0_5",
                  "C_mated_1", "C_mated_4", "C_mated_5", "C_mated_6", "C_mated_8", "C_mated_12", "C_mated_24")

DEgenes = read.delim("../out/DE_genes_shrunken1.3fold.txt", header = T, as.is = T)
load("../out/DESeq2_CvsV.RData")
ids = read.delim("../data/all_ids.txt", header = T, as.is = T)
case_grep = "mated"
control_grep = "virgin"
load("out/signature_matrix_broad.RData")

timepoints = c("0_5", "1", "2", "3", "4", "5", "6", "8", "12", "24")
```

Loop through time points:

```{r}
# Note: some genes causes a problem in cwFC evaluation: cw fold changes are all NA. Will remove these gene from DE gene list. 
DEgenes = DEgenes[,1]
DEgenes = DEgenes[! DEgenes %in% ids[ids[,1] %in% c("CG8929", "Baldspot", "CG32085"), 2]]

for(i in 7:length(timepoints)){
  # normalized counts:
  tp.cpm = cpm[,grep(timepoints[i], colnames(cpm))]
  
  # DE genes:
  tp = as.data.frame(CvsVresults[[1]][[i]])
  tp = tp[rownames(tp) %in% DEgenes,]
  tp = tp[tp$padj < 0.05,]
  tp = as.data.frame(cbind(rownames(tp), tp$padj, tp$log2FoldChange))
  rownames(tp) = tp[,1]
  colnames(tp) = c("Symbol", "Padj", "log2FC")
  tp$`Padj` = as.numeric(tp$`Padj`)
  tp$log2FC = as.numeric(tp$log2FC)
  
  # Change gene ids to symbols:
  idx = match(rownames(tp), ids[,2])
  rownames(tp) = ids[idx, 1]
  tp[,1] = rownames(tp)
  idx = match(rownames(tp.cpm), ids[,2])
  rownames(tp.cpm) = ids[idx, 1]
  tp.cpm = as.matrix(tp.cpm)

  # Calculation of cell-weighted fold changes:
  
  outdir = paste("scMappR_out", timepoints[i], sep = "_")
  
  toOut <- scMappR_and_pathway_analysis(count_file = tp.cpm, 
                                      signature_matrix = signature.matrix, 
                                      DEG_list = tp, 
                                      case_grep = case_grep,
                                      control_grep = control_grep, 
                                      #rda_path = "", 
                                      max_proportion_change = 10, 
                                        # Maximum cell-type proportion change – may be useful if there are many rare
                                        # cell-types. Alternatively, if a cell-type is only present in one condition but not the
                                        # other, it will prevent possible infinite or 0 cwFold-changes.
                                      print_plots = TRUE, 
                                      plot_names = "scMappR", 
                                      theSpecies = "dmelanogaster", 
                                      output_directory = outdir,
                                      sig_matrix_size = nrow(signature.matrix), 
                                      up_and_downregulated = TRUE, 
                                      internet = T, 
                                      toSave = TRUE, 
                                      path = outdir)
  
  # Evaluation of cw fold changes:
  evaluated <- cwFoldChange_evaluate(cwFC = toOut$cellWeighted_Foldchange$cellWeighted_Foldchange, 
                                   celltype_prop = toOut$cellWeighted_Foldchanges$cellType_Proportions, 
                                   DEG_list = tp)
  #save(toOut, file = paste(paste("cwFC_tp", timepoints[i], sep = "_"), "RData", sep = "."))
  save(evaluated, file = paste(paste("out/cwFC_evaluated_tp_head_only", timepoints[i], sep = "_"), "RData", sep = "."))
}
```

Rank-order of each gene in each celltype:

```{r}
head(evaluated$cwFoldchange_vs_bulk_rank_change)
```

Assignment of DEG to celltype:

```{r}
head(evaluated$cwFoldchange_gene_assigned)
```

# Assign genes to cell types where possible:

```{r}
load("out/cwFC_evaluated_tp_head_only_0_5.RData")
tp05 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_1.RData")
tp1 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_2.RData")
tp2 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_3.RData")
tp3 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_4.RData")
tp4 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_5.RData")
tp5 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_6.RData")
tp6 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_8.RData")
tp8 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_12.RData")
tp12 = evaluated$cwFoldchange_gene_assigned
load("out/cwFC_evaluated_tp_head_only_24.RData")
tp24 = evaluated$cwFoldchange_gene_assigned

l = list(tp05, tp1, tp2, tp3, tp4, tp5, tp6, tp8, tp12, tp24)

# make a list where each element contains all the info for a gene.
all_genes = c()
for(i in 1:length(l)){
  all_genes = unique(c(all_genes, unlist(lapply(l[[i]], function(x) names(x)))))
}
length(all_genes) # 391 genes

all_genes_res = list()
for(i in 1:length(all_genes)){
  gene = all_genes[i]
  all_res = c()
  for(j in 1:length(l)){
    res = lapply(l[[j]], function(x) gene %in% names(x))
    res = l[[j]][unlist(res)]
    res = lapply(res, function(x) x[names(x) == gene])
    res = cbind(names(res), res)
    all_res = rbind(all_res, res)
  }
  all_genes_res[[i]] = all_res
}
names(all_genes_res) = all_genes

# Keep for each gene cell type with max specificity (abs value), and set cutoff of 0.5:
s1 = lapply(all_genes_res, function(x) unlist(x[which.max(abs(as.numeric(x[,2]))),]))
names(s1) = names(all_genes_res)
check = lapply(s1, function(x) abs(as.numeric(x[2])) >= 0.5 )
s2 = s1[unlist(check)]
length(s2) # 367 genes

celltype = unlist(lapply(s2, function(x) x[1]))
gene = names(s2)
df = cbind(gene, celltype)
l = split(as.data.frame(df), df[,2])
ngenes = unlist(lapply(l, function(x) nrow(x)))
ngenes[order(ngenes)]
#neuron              muscle_cell               unanotated          epithelial_cell                 hemocyte 
#    7                        9                       14                       18                       19 
#male_reproductive_system           sensory_neuron               glial_cell                 fat_cell 
#                      29                       52                       58                      161
                      
save(df, file = "out/genes_assigned_to_celltype.RData")
```

